---
title: "Activity Monitoring"
author: "Duc"
date: "Sunday, January 10, 2016"
output: html_document
---

Set the default of echo to be true throughout the document:
```{r}
library(knitr)
opts_chunk$set(echo = TRUE)
```


### Load and check structure of data
```{r}
ds <- read.csv("activity.csv")
str(ds)
```

### What is mean total number of steps taken per day?
1. Calculate the total number of steps taken per day: we can do this by `tapply` function
```{r}
res1 <- tapply(ds$steps, ds$date, sum)
```

2. Histogram of the total number of steps taken each day
```{r}
hist(res1, xlab = "Total number of steps", ylab = "Number of days", 
     main = "Histogram of the total number of steps taken each day")
```

3. Mean and median of the total number of steps taken per day (ignoring NAs):
```{r}
mean(res1, na.rm = T)
median(res1, na.rm = T)
```

### What is the average daily activity pattern?
1. Time series plot of the average number of steps taken in each 5-minute interval, where we average over all days.
  1. compute the average for each 5-minute interval
```{r}
require(plyr)
intervals <- unique(ds$interval)
daily.ds <- ldply(intervals, function(ii) {
  avg.steps <- with(ds, mean(steps[interval == ii], na.rm = T))
  data.frame("interval" = ii, "avg.steps" = avg.steps)
})
```

  2. plot the time series
```{r}
require(ggplot2)
pp <- ggplot(data = daily.ds, aes(x = interval, y = avg.steps)) + geom_line(colour = "red")
pp <- pp + xlab("Interval") + ylab("Average number of steps taken")
pp
```

  3. Careful observation of the intervals (their ids actually) show that the ids >= 100 are actually not integers!! They actually have the form hhmm, for example 105 is 1 hour, 5 mins. Thus we need to parse the intervals using `lubridate` library.
```{r}
require(lubridate)
daily.ds <- transform(daily.ds, parse.interval = hm(interval/100))
```

  4. We found that the interval containing the maximum number of steps is 8:35AM. This is reasonable since that is the time we go to work.
```{r}
with(daily.ds, parse.interval[which.max(avg.steps)])
```

### Imputing missing values
1. The total number of missing values in the dataset is:
```{r}
sum(is.na(ds$steps))
```

2. We will impute missing values by the mean of the corresponding interval
```{r}
nas <- which(is.na(ds$steps))
full.steps <- numeric(nrow(ds))
full.steps[-nas] <- ds$steps[-nas]
full.steps[nas] <- sapply(nas, function(rr) {
  with(daily.ds, avg.steps[interval == ds$interval[rr]])
})
full.ds <- data.frame("steps" = full.steps, "date" = ymd(ds$date), "interval" = ds$interval)
```

3. After imputing missing values, the new histogram of the total number of steps taken each day:
```{r}
steps.per.day <- tapply(full.ds$steps, full.ds$date, sum)
par(mfrow = c(1, 2))
hist(steps.per.day, xlab = "Total number of steps", ylab = "Number of days", 
     main = "Histogram after imputing NAs", col = "blue")
hist(res1, xlab = "Total number of steps", ylab = "Number of days", 
     main = "Original histogram")
```

  1. Mean total number of steps:
```{r}
# before imputing missing values 
mean(res1, na.rm = T)
# after imputing missing values 
mean(steps.per.day)
```
The `mean` stays the same.

  2. Median total number of steps:
```{r}
# before imputing missing values 
median(res1, na.rm = T)
# after imputing missing values 
median(steps.per.day)
```
The `median` is slightly increase.

### Are there differences in activity patterns between weekdays and weekends?
1. Create indicators for weekdays and weekends:
```{r}
full.ds <- transform(full.ds, weekday = weekdays(date))
full.ds <- transform(full.ds, type = ifelse(weekday %in% c("Saturday", "Sunday"), "weekend", "weekday"))
```

2. 
```{r}
require(dplyr)
interval_full <- full.ds %>% group_by(interval, type) %>% summarise(avg.steps = mean(steps))
p2 <- ggplot(data = interval_full, aes(x = interval, y = avg.steps, colour = type)) + geom_line() + 
    facet_wrap(~ type, ncol = 1, nrow=2) + theme(legend.position="none")

p2 <- p2 + xlab("Interval") + ylab("Avg. number of steps taken")
p2
```


